---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pppower

<!-- badges: start -->
[![R-CMD-check](https://github.com/yiqunchen/pppower/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/yiqunchen/pppower/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of `pppower` is to conduct power analysis based on the Prediction-Powered Inference (PPI) framework. It is a convenient tools for planning and evaluating studies with PPI estimators.
The package provides analytical formulas, Monte Carlo estimators, plotting helpers, and sample-size solvers for both mean estimation and linear regression (PPI-OLS) workflows.

## Installation

You can install the development version of `pppower` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("yiqunchen/pppower")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(pppower)
## basic example code
set.seed(1)
```

**Start with a cross-fitted superpopulation data**

```{r superpop}
# Linear signal with coefficients 0.3, 0.8, ... and heavier Gaussian noise (sd = 2)
# Cross-fitted predictions (out-of-fold)
df <- simulate_crossfit_data(
  n = 3500, p = 5, family = gaussian(), #
  K = 5, seed = 20251007
)
str(df)
```

`simulate_crossfit_data()` builds a synthetic "superpopulation" with covariates, outcomes, and out-of-fold predictions (`fhat_cf`). These serve as inputs for both analytical formulas and Monte Carlo power simulations.

**Mean-estimation toolkit**

#### Analytical power and Monte Carlo calibration

```{r mean est}
theta_true <- mean(df$y)
theta0     <- theta_true - 0.35
delta      <- theta_true - theta0
var_f      <- var(df$fhat_cf)
var_res    <- var(df$y - df$fhat_cf)
N          <- 3000
n          <- 200
alpha      <- 0.05
sim        <- 2000

power_ppi_mean(delta, var_f, var_res, N, n, alpha)

simulate_power(
  delta   = delta,
  N       = N,
  n       = n,
  alpha   = alpha,
  var_f   = var_f,
  var_res = var_res,
  R       = sim
)
```

- `power_ppi_mean()` returns the exact normal-theory power.
- `simulate_power()` confirms the formula empirically for any variance inputs.

#### Monte Carlo with an explicit superpopulation

```{r mc}
mc_pp <- pppower:::simulate_power_ppi_mean(
  df      = df,
  N       = N,
  n       = n,
  alpha   = alpha,
  R       = sim,
  theta0  = theta0,
  seed    = 20251007
)
mc_pp$empirical_power
mc_pp$analytical_power
```

#### Sample-size planning

```{r sample size}
n_mean <- n_required_PP(
  delta   = delta,
  N       = N,
  alpha   = alpha,
  power   = 0.8,
  type    = "mean",
  var_f   = var_f,
  var_res = var_res
)
n_mean
```

`n_required_PP()` now supports the PP mean, PP-OLS, and custom variance decompositions through the type switch (logistic regression in the future).

**Linear Regression**

```{r fit ols}
fit_ols <- ppi_ols_fit(
  y      = "y",
  f_col  = "fhat_cf",
  formula = ~ x1 + x2 + x3 + x4 + x5,
  data_l = df[1:n, ],
  data_u = df[(n + 1):(n + N), ]
)

summary(fit_ols)

pieces      <- fit_ols$pieces
coef_names  <- colnames(pieces$V_u)       # matches V_u/V_l dimension
c_vec       <- numeric(length(coef_names))
names(c_vec) <- coef_names
c_vec["x1"] <- 1

delta_ols <- sum(c_vec * fit_ols$coef)
```

`ppi_ols_fit()` returns coefficient estimates, sandwich covariance, and the $V_u$, $V_l$ components required for power calculations.

`ppi_ols_wald()` offers Wald tests for arbitrary contrasts.

#### Analytical and Monte Carlo power

```{r power compare}
# superpopulation contrast for the given c
X_full      <- model.matrix(~ x1 + x2 + x3 + x4 + x5, df)
beta_true   <- pppower:::ols_fit(X_full, df$y)$coef
contrast_true <- sum(c_vec * beta_true)

# the effect size you want to study
delta_target <- delta_ols         # if you want to stick with the sample estimate
theta0       <- contrast_true - delta_target

pppower:::power_ppi_ols(
  delta = delta_target,
  V_u   = pieces$V_u,
  V_l   = pieces$V_l,
  N     = pieces$N,
  n     = pieces$n,
  c     = c_vec,
  alpha = alpha
)

pppower:::simulate_power_ppi_ols(
  df      = df,
  formula = ~ x1 + x2 + x3 + x4 + x5,
  N       = pieces$N,
  n       = pieces$n,
  c       = c_vec,
  theta0  = theta0,
  alpha   = alpha,
  R       = sim,
  seed    = 20251007
)$empirical_power
```

#### Sample-size planning for PP-OLS

```{r n required ols}
n_ols <- n_required_PP(
  delta = delta_ols,
  N     = pieces$N,
  alpha = alpha,
  power = 0.8,
  type  = "ols",
  V_u   = pieces$V_u,
  V_l   = pieces$V_l,
  c     = c_vec
)
n_ols
```

#### Power Curves

**Mean Estimation**

```{r, dpi = 200}
effect_grid <- seq(-0.4, 0.4, by = 0.05)
curve_fixed <- type1_error_curve_mean(
  effect_grid = effect_grid,
  N = N,
  n = n,
  var_f = var_f,
  var_res = var_res,
  alpha = alpha,
  R = sim,
  seed = 20251007
)

plot_type1_error_curve(
  curve_fixed,
  main = "Power curve from variance inputs"
)
```

**Cross-fitted Data Generation Processes**

```{r dgp-power-curve, dpi = 200}
curve_dgp <- type1_error_curve_mean_dgp(
  effect_grid = effect_grid,
  N = N,
  n = n,
  family = stats::gaussian(),
  superpop_n = 10000,
  p = 5,
  K = 5,
  alpha = alpha,
  R = sim,
  seed = 20251007
)

plot_type1_error_curve(
  curve_dgp,
  main = "Power curve with cross-fitted DGP (Gaussian)"
)
```

`plot_type1_error_curve()` is shared by both outputs and lets you overlay empirical/analytical curves and nominal reference lines.


**Reference of key functions**

  - Simulation:
  `simulate_crossfit_data()`: synthetic GLM data with cross-fitted predictions.

  `simulate_power()` / `simulate_power_ppi_mean()` / `simulate_power_ppi_ols()`: Monte Carlo power.

  - Analytical power:
  `power_ppi_mean()`: PP mean estimator.
  `power_ppi_ols()`: Linear contrasts in PP-OLS.

  - Estimation & inference:
  `ppi_ols()` (*currently internal*), `ppi_ols_fit()`, `ppi_ols_wald()` (*currently internal*).

  - Planning:
  `n_required_PP()`: unified sample-size solver (type = "mean", "ols", or "custom").

  - Plotting:
  `power_curve_mean()`, `power_curve_mean_dgp()`, `power_curve_gaussian()`, `power_curve_binomial()`.
  `type1_error_curve_mean()`, `type1_error_curve_mean_dgp()`, `plot_type1_error_curve()`.

  Each example above is self-contained; adjust Monte Carlo replicates (R) upward for production studies.

**Contributing**
  - Issues and pull requests are welcome. Please add tests under tests/testthat/ and run `devtools::test()` before submitting changes.


```{r session-info, echo = FALSE}
devtools::session_info()
```