---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# pppower

<!-- badges: start -->
[![R-CMD-check](https://github.com/yiqunchen/pppower/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/yiqunchen/pppower/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

The goal of `pppower` is to conduct power analysis based on the Prediction-Powered Inference (PPI) framework. It is a convenient tools for planning and evaluating studies with PPI estimators.
The package provides analytical formulas, Monte Carlo estimators, plotting helpers, and sample-size solvers for both mean estimation and linear regression (PPI-OLS) workflows.

## Installation

You can install the development version of `pppower` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("yiqunchen/pppower")
```

## Vignettes

- **Prediction-Powered Power Planning**  

Open in R:  

``` r
vignette("intro-ppi", package = "pppower")
```

```{r session-info, echo = FALSE}
devtools::session_info()
```