---
title: "Prediction-Powered Power Planning"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Prediction-Powered Power Planning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-chunk, include = FALSE}
knitr::opts_chunk$set(
  fig.path = "figures/intro-ppi-",
  fig.width = 6, fig.height = 4,
  collapse = TRUE, comment = "#>"
)
```

## Overview

`pppower` provides analytical and Monte Carlo tools for **prediction-powered inference (PPI)**. We cover two main workflows:

* **Prediction-powered mean estimator** - the target parameter is the population mean \(\theta = \mathbb{E}[Y]\) corrected via predictions \(f(X)\).
* **Prediction-powered linear regression (PPI-OLS)** - the estimand is a linear contrast \(c^\top \beta\) from a PPI-adjusted OLS estimator.

---

## 1. Prediction-Powered Mean Estimation

### Notation

Notation

We test the null hypothesis H_0: \theta = \theta_0
versus a two-sided alternative, where:
-	\delta = \theta - \theta_0: effect size of interest
-	N: unlabeled sample size (used to estimate \mathbb{E}[f(X)])
-	n: labeled sample size (used to correct prediction error)
-	\sigma_f^2 = \mathrm{Var}(f(X)): variance of the model prediction
-	\sigma_{\text{res}}^2 = \mathrm{Var}(Y - f(X)): residual variance after prediction

These quantities are typically estimated from prior data, pilot studies, or external models — but for planning and power analysis, we can specify them directly.


```{r mean-inputs}
library(pppower)

# Study design parameters
N     <- 2000     # unlabeled sample size
n     <- 200      # labeled sample size
alpha <- 0.05     # two-sided significance level
sim   <- 2000     # Monte Carlo replicates

# Population-level quantities (from prior knowledge, pilot data, or assumptions)
theta0   <- 0.0   # null value of the mean
delta    <- 0.15  # effect size (true mean - theta0)
var_f    <- 0.45  # variance of f(X)
var_res  <- 0.80  # residual variance
```

### Analytical power

With these inputs, we can calculate the analytical power of the PPI mean estimator using the closed-form normal approximation:

`power_ppi_mean()` implements the normal-theory power formula, using the standard error \(\sqrt{\sigma_f^2 / N + \sigma_{\text{res}}^2 / n}\).

```{r mean-analytic}
power_ppi_mean(
  delta = delta,
  var_f = var_f,
  var_res = var_res,
  N = N,
  n = n,
  alpha = alpha
)
```

### Monte Carlo calibration (parametric)

`simulate_power()` mimics the null distribution with Monte Carlo samples of the test statistic.

```{r mean-simulate}
simulate_power(
  delta = delta,
  N = N,
  n = n,
  alpha = alpha,
  var_f = var_f,
  var_res = var_res,
  R = sim
)
```

### Monte Carlo calibration (resampling superpopulation)

`pppower:::simulate_power_ppi_mean()` repeatedly resamples labeled/unlabeled splits from the synthetic population, recomputing the PPI estimator each time.

```{r mean-pp}
moments_ppi <- list(
  delta   = delta,
  var_f   = var_f,
  var_res = var_res
)

# Run the full Monte Carlo simulation and compare with analytical power
mc_pp <- pppower:::simulate_power_ppi_mean(
  R = sim,
  n = n,
  N = N,
  alpha = alpha,
  family = gaussian(),
  moments = moments_ppi,
  theta0 = theta0,
  seed = 2024
)

# Results: empirical rejection rate (Monte Carlo) vs. analytical formula
mc_pp$empirical_power
mc_pp$analytical_power
```

### PPI++ Mean Estimation

```{r ppiplusplus, eval = FALSE, include = FALSE}
moments_ppiplus <- list(
  delta     = 0.15,
  sigma_y2  = 1.10,     # total variance of Y
  sigma_f2  = 0.45,     # variance of f(X)
  cov_y_f   = 0.45,     # Cov(Y, f(X)), equals var_f if unbiased
  var_f     = 0.45,
  var_res   = 0.65      # residual variance
)

# Analytical power based on population moments and optimal lambda
analytical_ppiplus <- power_ppi_pp_mean(
  delta      = moments_ppiplus$delta,
  N          = N,
  n          = n,
  sigma_y2   = moments_ppiplus$sigma_y2,
  sigma_f2   = moments_ppiplus$sigma_f2,
  cov_y_f    = moments_ppiplus$cov_y_f,
  var_f      = moments_ppiplus$var_f,
  var_res    = moments_ppiplus$var_res,
  alpha      = alpha
)
analytical_ppiplus

# Monte Carlo calibration (repeated simulation from the population)
set.seed(20240623L)
ppplusplus_mc <- pppower:::simulate_power_ppiplus_mean(
  R          = 2000,
  n          = n,
  N          = N,
  alpha      = alpha,
  family     = gaussian(),
  moments    = moments_ppiplus,
  lambda_type = "plugin",
  seed       = 20240623L
)

# Compare analytical and empirical power
ppplusplus_mc$analytical_power
ppplusplus_mc$empirical_power

# Inspect lambda used and average across simulations
ppplusplus_mc$avg_lambda
```


### PPI/PPI++ OLS

```{r PPI OLS}
c_list <- list(
  c1 = c(1, 0, 0),        # intercept
  c2 = c(0, 1, 0),        # slope on x1
  c3 = c(0, 0, 1),        # slope on x2
  c4 = c(0, 1, -1),       # difference x1 - x2
  c5 = c(0, 0.5, 0.5)     # average (x1 + x2)/2
)

beta_true <- c(0.7, 0.4, -0.3)
p <- length(beta_true)

# Choose a null for each contrast
theta0_map <- c(
  c1 = 0.0,
  c2 = 0.25,
  c3 = 0.0,
  c4 = 0.0,
  c5 = 0.0
)

# Variance components and matrices
sigma_y2_pop <- 1.0
sigma_f2_pop <- 0.45

H_pop         <- diag(p)
Sigma_YY_pop  <- sigma_y2_pop * H_pop
Sigma_ff_pop  <- sigma_f2_pop * H_pop
Sigma_Yf_pop  <- matrix(0, nrow = p, ncol = p)

set.seed(20240625L)

run_one_contrast <- function(name, c_vec, theta0) {
  delta_c <- as.numeric(c(c_vec %*% beta_true) - theta0)

  moments_ols <- list(
    delta    = delta_c,
    H        = H_pop,
    Sigma_YY = Sigma_YY_pop,
    Sigma_ff = Sigma_ff_pop,
    Sigma_Yf = Sigma_Yf_pop
  )

  res <- pppower:::simulate_power_ppi_ols(
    R        = sim,
    n        = n,
    N        = N,
    contrast = c_vec,
    alpha    = alpha,
    family   = gaussian(),
    moments  = moments_ols,
    seed     = sample.int(.Machine$integer.max, 1)
  )

  mc_se <- sqrt(res$analytical_power * (1 - res$analytical_power) / sim)
  tol   <- max(3 * mc_se, 0.01)
  diff  <- abs(res$empirical_power - res$analytical_power)

  data.frame(
    contrast   = name,
    theta0     = theta0,
    delta      = delta_c,
    analytical = res$analytical_power,
    empirical  = res$empirical_power,
    abs_diff   = diff,
    tol        = tol,
    pass       = diff < tol,
    stringsAsFactors = FALSE
  )
}

res_list <- mapply(
  run_one_contrast,
  name  = names(c_list),
  c_vec = c_list,
  theta0 = theta0_map[names(c_list)],
  SIMPLIFY = FALSE
)

ols_multi <- do.call(rbind, res_list)
ols_multi
```

```{r PPI++ OLS}
c_list <- list(
  c1 = c(1, 0, 0),        
  c2 = c(0, 1, 0),        
  c3 = c(0, 0, 1),        
  c4 = c(0, 1, -1),       
  c5 = c(0, 0.5, 0.5)     
)

beta_true <- c(0.7, 0.4, -0.3)
p <- length(beta_true)

theta0_map <- c(
  c1 = 0.0,
  c2 = 0.25,
  c3 = 0.0,
  c4 = 0.0,
  c5 = 0.0
)

sigma_y2_pop <- 1.0
sigma_f2_pop <- 0.45

H_pop         <- diag(p)
Sigma_YY_pop  <- sigma_y2_pop * H_pop
Sigma_ff_L_pop <- sigma_f2_pop * H_pop
Sigma_ff_U_pop <- sigma_f2_pop * H_pop
Sigma_Yf_pop  <- sigma_f2_pop * H_pop

set.seed(20240626L)

run_one_contrast_ppiplus <- function(name, c_vec, theta0) {
  delta_c <- as.numeric(c(c_vec %*% beta_true) - theta0)

  moments_ppiplus <- list(
    delta      = delta_c,
    H_L        = H_pop,
    H_U        = H_pop,
    Sigma_YY   = Sigma_YY_pop,
    Sigma_ff_l = Sigma_ff_L_pop,
    Sigma_ff_u = Sigma_ff_U_pop,
    Sigma_Yf   = Sigma_Yf_pop
  )

  res <- pppower:::simulate_power_ppi_pp_ols(
    R        = sim,
    n        = n,
    N        = N,
    contrast = c_vec,
    alpha    = alpha,
    family   = gaussian(),
    lambda_type = "plugin",
    moments  = moments_ppiplus,
    seed     = sample.int(.Machine$integer.max, 1)
  )

  mc_se <- sqrt(res$analytical_power * (1 - res$analytical_power) / sim)
  tol   <- max(3 * mc_se, 0.01)
  diff  <- abs(res$empirical_power - res$analytical_power)

  data.frame(
    contrast   = name,
    theta0     = theta0,
    delta      = delta_c,
    analytical = res$analytical_power,
    empirical  = res$empirical_power,
    abs_diff   = diff,
    tol        = tol,
    pass       = diff < tol,
    avg_lambda = res$avg_lambda,
    stringsAsFactors = FALSE
  )
}

res_list_ppiplus <- mapply(
  run_one_contrast_ppiplus,
  name  = names(c_list),
  c_vec = c_list,
  theta0 = theta0_map[names(c_list)],
  SIMPLIFY = FALSE
)

ols_ppiplus_multi <- do.call(rbind, res_list_ppiplus)
ols_ppiplus_multi

```

### Sample-size planning

`n_required_pp()` solves for the labeled sample size that meets a power target.

```{r mean-n}
n_required_pp(
  delta = delta,
  N = N,
  alpha = alpha,
  power = 0.80,
  type = "mean",
  var_f = var_f,
  var_res = var_res
)
```

### Power and Type I error curves

`power_curve_mean()` sweeps over candidate labeled sample sizes, while `plot_type1_error_curve()` draws empirical vs. analytical rejection probabilities.

```{r mean-power-curve}
curve_fixed <- pppower:::power_curve_mean(
  n_grid = seq(50, 250, by = 50),
  delta = delta,
  N = N,
  var_f = var_f,
  var_res = var_res,
  alpha = alpha,
  R = sim
)

curve_fixed

plot(
  curve_fixed$n,
  curve_fixed$power_empirical,
  type = "b",
  pch = 19,
  col = "steelblue",
  xlab = "Labeled sample size n",
  ylab = "Power",
  main = "Power curve (mean estimator)"
)
lines(curve_fixed$n, curve_fixed$power_exact, col = "firebrick", lwd = 2, lty = 2)
legend("bottomright", legend = c("Empirical", "Analytical"),
       col = c("steelblue", "firebrick"), lty = c(1, 2), pch = c(19, NA), lwd = c(1, 2), bty = "n")
```

To study Type I error and power as a function of \(\delta\), use `type1_error_curve_mean()` or, for a fully simulated DGP, `type1_error_curve_mean_dgp()`:

```{r mean-type1, eval = TRUE}
curve_parametric <- type1_error_curve_mean_dgp(
  effect_grid = seq(-0.5, 0.5, by = 0.05),
  N = N,
  n = n,
  var_f = var_f,
  var_res = var_res,
  theta = 1.0,
  alpha = 0.05,
  R = 2000,
  seed = 20240627L
)

pppower:::plot_type1_error_curve(
  curve_parametric,
  main = "Type I Error / Power Curve (Parametric, No Superpopulation)"
)
```

---

## 2. Prediction-Powered OLS (PPI-OLS)

### Notation

Let \(\beta\) be the PPI-adjusted coefficient vector and \(c\) the contrast of interest.

- Estimand: \(c^\top \beta\).
- Variance: \(c^\top (V_u / N + V_l / n) c\), where `V_u`/`V_l` are per-observation sandwich covariances for the unlabeled (imputation) and labeled (rectifier) regressions.

### OLS set up

```{r ols-setup}
# Study design
N     <- 2000      # unlabeled sample size
n     <- 200       # labeled sample size
alpha <- 0.05
sim   <- 2000

# True coefficient vector (assumed or estimated from prior data)
beta_true <- c(0.7, 0.4, -0.3, 0.2)

# Contrast vector for hypothesis test
c_vec <- c(0, 1, 0, 0)      # test second coefficient

# Effect size (difference from null)
theta0  <- 0.25
delta   <- as.numeric(t(c_vec) %*% beta_true - theta0)

# Variance components (assumed known from pilot data or prior studies)
sigma_y2 <- 1.0       # total variance of Y
sigma_f2 <- 0.45      # variance explained by predictor model

# Per-observation covariance matrices
p <- length(beta_true)
H_pop         <- diag(p)                       # design matrix Hessian
Sigma_YY_pop  <- sigma_y2 * H_pop              # Var(Y)
Sigma_ff_pop  <- sigma_f2 * H_pop              # Var(f(X))
Sigma_Yf_pop  <- sigma_f2 * H_pop              # Cov(Y, f(X))
```

### Analytical Power

We can compute the analytical power for the PPI-OLS estimator using these components:

```{r ols-fit}
power_ppi_ols(
  delta = delta,
  V_u   = Sigma_ff_pop,
  V_l   = Sigma_YY_pop,
  N     = N,
  n     = n,
  c     = c_vec,
  alpha = alpha
)
```


### Monte Carlo calibration

Suppose we focus on `x1`:

```{r ols-mc}
moments_ols <- list(
  delta    = delta,
  H        = H_pop,
  Sigma_YY = Sigma_YY_pop
)

ols_mc <- pppower:::simulate_power_ppi_ols(
  R        = sim,
  n        = n,
  N        = N,
  contrast = c_vec,
  alpha    = alpha,
  family   = gaussian(),
  moments  = moments_ols,
  seed     = 20240625L
)

ols_mc$analytical_power
ols_mc$empirical_power
abs(ols_mc$empirical_power - ols_mc$analytical_power)
```

### Sample-size solver

```{r ols-n}

target <- 0.90

n_required_pp(
  delta = delta,
  N = N,
  alpha = alpha,
  power = target,
  type = "ols",
  V_u = Sigma_ff_pop,
  V_l = Sigma_YY_pop,
  c = c_vec
)

n_required_ppi_pp(
  delta      = delta,
  N          = N,
  alpha      = alpha,
  power      = target,
  type       = "ols",
  lambda_mode = "oracle",
  c          = c_vec,
  H_L        = H_pop,
  H_U        = H_pop,
  Sigma_YY   = Sigma_YY_pop,
  Sigma_ff_l = Sigma_ff_pop,
  Sigma_ff_u = Sigma_ff_pop,
  Sigma_Yf   = Sigma_Yf_pop
)

```

---

## Summary Table

| Task                                          | Functions                                                            |
|-----------------------------------------------|-----------------------------------------------------------------------|
| Generate cross-fitted superpopulation         | `simulate_crossfit_data()`                                           |
| Mean estimator power & calibration            | `power_ppi_mean()`, `simulate_power()`, `pppower:::simulate_power_ppi_mean()` |
| Mean estimator sample size                    | `n_required_pp(type = "mean", ...)`                                   |
| Mean estimator curves                         | `power_curve_mean()`, `type1_error_curve_mean[_dgp]()`, `plot_type1_error_curve()` |
| Fit prediction-powered OLS                    | `ppi_ols_fit()`, `summary()`                                          |
| OLS contrasts and power                       | `pppower:::ppi_ols_wald()`, `power_ppi_ols()`, `pppower:::simulate_power_ppi_ols()`       |
| OLS sample size                               | `n_required_pp(type = "ols", ...)`                                    |

Together these tools help you explore prediction quality, diagnose finite-sample effects, and plan labeled sample sizes for both mean and regression targets.
