---
title: "Quickstart: Prediction-Powered Power & Sample Size"
output:
  rmarkdown::html_document:
    toc: false
vignette: >
  %\VignetteIndexEntry{Quickstart: Prediction-Powered Power & Sample Size}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-chunk, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## What this vignette covers

This is the **basic-use** guide for `pppower`. It shows the minimum calls you
need for power, sample size, and paired designs. See the “Detailed Dive” and
“Real Data” tabs on the pkgdown site for derivations and LLM-as-a-judge setups.

```{r}
library(pppower)
```

### What it does (analogy to `pwr`)

- Drop-in power / sample-size calculators that leverage predictions on unlabeled data (PPI++).
- Same spirit as `pwr::pwr.t.test`, but variance comes from predictions + residuals.
- Works for means / paired differences and binary surrogates (LLM judge).

### 1. Power for a prediction-powered mean test

```{r}
power_ppi_mean(
  delta    = 0.2,   # effect size
  N        = 5000,  # unlabeled size
  n        = 200,   # labeled size
  var_f    = 0.6,   # Var(f(X))
  var_res  = 0.4,   # Var(Y - f(X))
  cov_y_f  = 0.6    # Cov(Y, f)
)
```

### 2. Required labeled size for target power

```{r}
power_ppi_mean(
  delta    = 0.2,
  N        = 5000,
  n        = NULL,
  power    = 0.80,
  sigma_y2 = 1.0,
  sigma_f2 = 0.6,
  cov_y_f  = 0.6
)
```

You can also supply prediction metrics (MSE, $R^2$, sensitivity/specificity)
instead of raw moments; see the “Detailed Dive” vignette.

### 3. Paired designs in one line

```{r}
power_ppi_paired(
  delta    = 0.3,   # mean difference
  N        = 1000,  # unlabeled pairs
  n        = 50,    # labeled pairs
  sigma_D2 = 1.0,
  rho_D    = 0.7
)
```

### Where next?

- **Detailed Dive:** formulas, metrics-to-moments mapping, regression contrasts.
- **Real Data:** LLM-as-a-judge binary surrogates (EIF), required calibration size.
