---
title: "Prediction-Powered Sample Size Calculation"
subtitle: "Using `n_required_ppi_pp()` for Mean and Regression PPI/PPI++"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Prediction-Powered Sample Size Calculation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.path = "figures/ppi-sample-size-",
  fig.width = 10, fig.height = 8,
  collapse = TRUE, comment = "#>"
)
library(pppower)
library(ggplot2)
library(dplyr)
library(tidyr)

set.seed(20251129)
```


## Introduction

Prediction-Powered Inference (PPI) and PPI++ reduce labeled sample size needs by using a prediction model trained on large unlabeled datasets.

This vignette demonstrates how to compute required **required labeled sample size** under:

- Mean Estimation
- OLS and GLM recression contrasts
- Direct moments inputs or prediction metrics (confusion matrix, brier score, MSE, $R^2$, etc.)
- Vanilla (PPI), Oracle, and user-specified $\lambda$ (PPI++)

All examples use the main function: `n_required_ppi_pp()` which solves for the smallest labeled sample size $(n)$ such that the asymptotic variance meets the target power level.


## Overview of the Sample Size Function

The core function has the following interface:
```{r overview, eval = FALSE}
n_required_ppi_pp(
  delta, N,
  type       = c("mean", "regression"),
  lambda_mode = c("vanilla", "oracle", "user"),
  lambda_user = NULL,
  # mean inputs
  sigma_y2, sigma_f2, cov_y_f,
  var_f, var_res,
  metrics, metric_type,
  # regression inputs
  c, H_L, H_U,
  Sigma_YY, Sigma_ff_l, Sigma_ff_u, Sigma_Yf,
  mode = c("error", "cap")
)
```

Explanation:

- `delta`: the effect size $(c^\top \beta - c^\top\beta_0)$ for regression (OLS/GLM), or a shift in the mean.
- `N`: size of the unlabeled dataset, typically very large.
- `lambda_mode`: determines whether $\lambda$ is fixed at 1 (Vanilla PPI estimator), chosen by the user (`"user"`), or optimized (`"oracle"`).
    - For mean mode, you may provide moments (`sigma_y2`, `sigma_f2`, `cov_y_f`) or prediction metrics (`mse`, `var_f`, accuracy, etc.).
	- For regression mode, you pass Hessians and covariance blocks computed by `compute_ppi_blocks()`.


## Mean-Mode Sample Size — Direct Moments

In the simplest case, you directly supply the population moments:

- $\sigma_y^2 = \mathrm{Var}(Y)$,
- $\sigma_f^2 = \mathrm{Var}(f(X))$,
- $\mathrm{Cov}(Y,f)$

```{r mean-direct}
delta  <- 0.20
N      <- 3000
alpha  <- 0.05
power  <- 0.80

sigma_y2 <- 1.00
sigma_f2 <- 0.40
cov_y_f  <- 0.15

n_required_ppi_pp(
  delta      = delta,
  N          = N,
  type       = "mean",
  lambda_mode = "vanilla",
  sigma_y2   = sigma_y2,
  sigma_f2   = sigma_f2,
  cov_y_f    = cov_y_f
)

n_required_ppi_pp(
  delta      = delta,
  N          = N,
  type       = "mean",
  lambda_mode = "oracle",
  sigma_y2   = sigma_y2,
  sigma_f2   = sigma_f2,
  cov_y_f    = cov_y_f
)
```

The stronger the correlation between outcome ($Y$) and the model prediction ($f(X)$), the more sample size reduction PPI++ yields relative to vanilla PPI.

## Mean-Mode Sample Size — Using Prediction Metrics

Instead of providing moments, you can supply **predictive performance metrics**.
```{r mean-metrics}
metrics <- list(
  type    = "continuous",
  var_y   = 1.00,   # total outcome variance
  mse     = 0.20,   # very strong model (small residual error)
  var_f   = 0.80,   # predictions have high variance
  cov_y_f = 0.70,   # very high correlation model
  m_obs   = 500
)

n_required_ppi_pp(
  delta       = 0.15,
  N           = 520,
  type        = "mean",
  lambda_mode = "vanilla",
  metrics     = metrics,
  metric_type = "continuous"
)

n_required_ppi_pp(
  delta       = 0.15,
  N           = 520,
  type        = "mean",
  lambda_mode = "oracle",
  metrics     = metrics,
  metric_type = "continuous"
)

```

`m_obs` is the number of labeled samples used to estimate the prediction metrics (MSE, variance, accuracy, etc.).
Finite-sample corrections ensure the derived moments remain unbiased / feasible.


## Mean-Mode Sample Size — Using Confusion matrix

You can also supply classifier performance through a confusion matrix:

```{r mean-metrics-conf}
metrics <- list(
  type = "classification",
  # Confusion matrix (strong classifier)
  tp = 225,   # true positives
  fp =  25,   # false positives
  fn =  25,   # false negatives
  tn = 225,   # true negatives

  # Total observed labeled data used to compute the metrics
  m_obs = 500
)

# For reference:
# p_y = (tp + fn) / m_obs = (225 + 25)/500 = 0.50  (balanced)
# accuracy = (tp + tn)/500 = 450/500 = 0.90

# PPI sample-size calculation, λ=1
n_required_ppi_pp(
  delta       = 0.15,
  N           = 520,
  type        = "mean",
  lambda_mode = "vanilla",
  metrics     = metrics,
  metric_type = "classification"
)

# PPI++ oracle-λ
n_required_ppi_pp(
  delta       = 0.15,
  N           = 520,
  type        = "mean",
  lambda_mode = "oracle",
  metrics     = metrics,
  metric_type = "classification"
)

```

For binary outcomes, the function internally computes:

- prevalence ($p_y$)
- prediction prevalence ($p_f$)
- variance of residuals
- covariance ( $\mathrm{Cov}(Y,f)$ )

using identities derived from the confusion matrix.


## Regression-Mode Sample Size — OLS

We now demonstrate the **regression contrast** mode.

Generate Data: 

```{r ols-data}
p    <- 4
n_l  <- 400
n_u  <- 3000
beta <- c(0.7, -0.5, 0.2, 1.0)

X_l <- matrix(rnorm(n_l*p), n_l, p)
X_u <- matrix(rnorm(n_u*p), n_u, p)

Y_l <- drop(X_l %*% beta + rnorm(n_l, sd = 0.7))
f_l <- drop(X_l %*% beta + rnorm(n_l, sd = 0.6))
f_u <- drop(X_u %*% beta + rnorm(n_u, sd = 0.6))
```

Construct PPI++ blocks for sandwich variance estimators:

```{r ols-blocks}
blocks <- compute_ppi_blocks(
  model_type = "ols",
  X_l = X_l, Y_l = Y_l, f_l = f_l,
  X_u = X_u, f_u = f_u,
  beta = beta
)
```

Solve sample size for contrast $c^\top \beta$:

```{r ols-contrast}
c_vec <- c(0, 1, 0, 0)
delta <- as.numeric(t(c_vec) %*% beta)

n_required_ppi_pp(
  delta       = delta,
  N           = n_u,
  power       = 0.90,
  type        = "regression",
  lambda_mode = "vanilla",
  c           = c_vec,
  H_L         = blocks$H_L,
  H_U         = blocks$H_U,
  Sigma_YY    = blocks$Sigma_YY,
  Sigma_ff_l  = blocks$Sigma_ff_l,
  Sigma_ff_u  = blocks$Sigma_ff_u,
  Sigma_Yf    = blocks$Sigma_Yf
)

n_required_ppi_pp(
  delta       = delta,
  N           = n_u,
  power       = 0.90,
  type        = "regression",
  lambda_mode = "oracle",
  c           = c_vec,
  H_L         = blocks$H_L,
  H_U         = blocks$H_U,
  Sigma_YY    = blocks$Sigma_YY,
  Sigma_ff_l  = blocks$Sigma_ff_l,
  Sigma_ff_u  = blocks$Sigma_ff_u,
  Sigma_Yf    = blocks$Sigma_Yf
)

```

## Regression-Mode Sample Size — GLM: Logistic Regression

(We follow similar steps as described in OLS case above)

```{r glm-data}
p    <- 3
n_l  <- 500
n_u  <- 2500
beta <- c(1.2, -0.6, 0.5)

X_l <- matrix(rnorm(n_l*p), n_l, p)
X_u <- matrix(rnorm(n_u*p), n_u, p)

eta_l <- drop(X_l %*% beta)
eta_u <- drop(X_u %*% beta)

prob_l <- plogis(eta_l)
prob_u <- plogis(eta_u)

Y_l <- rbinom(n_l, 1, prob_l)

f_l <- plogis(eta_l + rnorm(n_l, sd = 1.5))
f_u <- plogis(eta_u + rnorm(n_u, sd = 1.5))
```


Construct PPI++ blocks for sandwich variance estimators similar to previous OLS case:

```{r glm-block}
blocks_glm <- compute_ppi_blocks(
  model_type = "glm",
  family     = "binomial",
  X_l = X_l, Y_l = Y_l, f_l = f_l,
  X_u = X_u, f_u = f_u,
  beta = beta
)
```

Then solve for sample size:

```{r glm-contrast}
c_vec <- c(1, 0, 0)
delta <- as.numeric(t(c_vec) %*% beta)

n_required_ppi_pp(
  delta       = delta,
  N           = n_u,
  type        = "regression",
  lambda_mode = "vanilla",
  c           = c_vec,
  H_L         = blocks_glm$H_L,
  H_U         = blocks_glm$H_U,
  Sigma_YY    = blocks_glm$Sigma_YY,
  Sigma_ff_l  = blocks_glm$Sigma_ff_l,
  Sigma_ff_u  = blocks_glm$Sigma_ff_u,
  Sigma_Yf    = blocks_glm$Sigma_Yf
)

n_required_ppi_pp(
  delta       = delta,
  N           = n_u,
  type        = "regression",
  lambda_mode = "oracle",
  c           = c_vec,
  H_L         = blocks_glm$H_L,
  H_U         = blocks_glm$H_U,
  Sigma_YY    = blocks_glm$Sigma_YY,
  Sigma_ff_l  = blocks_glm$Sigma_ff_l,
  Sigma_ff_u  = blocks_glm$Sigma_ff_u,
  Sigma_Yf    = blocks_glm$Sigma_Yf
)
```

## Handling infeasible cases 

If the required labeled ($n$) exceeds the unlabeled ($N$), the function throws an error by default (`mode = "error"`):

```{r infeasible, error=TRUE, message=TRUE}
out <- n_required_ppi_pp(
  delta = 0.03,
  N     = 1000,
  type  = "mean",
  sigma_y2 = 1,
  sigma_f2 = 0.2,
  cov_y_f  = 0.05,
  mode = "error"
)
```

Switch to `mode = "cap"` to get an estimate of power at $n = N$:

```{r infeasible-cap}
out <- n_required_ppi_pp(
  delta = 0.03,
  N     = 1000,
  type  = "mean",
  sigma_y2 = 1,
  sigma_f2 = 0.2,
  cov_y_f  = 0.05,
  mode = "cap"
)

out
attr(out, "achieved_power")
```


## Power Curve Visualization

This chunk visualizes how required ($n$) decreases as effect size grows, and how oracle $\lambda$ for PPI++ estimator provides a clear reduction in sample size estimations over vanilla PPI:

```{r power-curve-sample-size, warning=FALSE, fig.align="center"}
delta_grid <- seq(0.10, 0.40, length.out = 40)

## Shared parameters (mean-estimation case)
N       <- 3000
sy2     <- 1.0      # Var(Y)
sf2     <- 0.4      # Var(f)
cyf     <- 0.15     # Cov(Y, f)
lambda_user_value <- 0.70   # example user-specified λ

## Compute required n for each λ mode
results <- lapply(delta_grid, function(d) {
  data.frame(
    delta = d,
    vanilla = n_required_ppi_pp(
      delta       = d,
      N           = N,
      type        = "mean",
      lambda_mode = "vanilla",
      sigma_y2    = sy2,
      sigma_f2    = sf2,
      cov_y_f     = cyf
    ),
    oracle = n_required_ppi_pp(
      delta       = d,
      N           = N,
      type        = "mean",
      lambda_mode = "oracle",
      sigma_y2    = sy2,
      sigma_f2    = sf2,
      cov_y_f     = cyf
    ),
    user = n_required_ppi_pp(
      delta       = d,
      N           = N,
      type        = "mean",
      lambda_mode = "user",
      lambda_user = lambda_user_value,
      sigma_y2    = sy2,
      sigma_f2    = sf2,
      cov_y_f     = cyf
    )
  )
})

df <- bind_rows(results) |>
  pivot_longer(cols = c("vanilla", "oracle", "user"),
               names_to = "mode",
               values_to = "n")

ggplot(df, aes(delta, n, color = mode)) +
  geom_line(size = 1.2) +
  scale_color_manual(
    values = c(
      vanilla = "#1B9E77",
      oracle  = "#D95F02",
      user    = "#7570B3"
    ),
    labels = c(
      vanilla = "PPI (vanilla λ = 1)",
      oracle  = "PPI++ (oracle λ*)",
      user    = paste0("User λ = ", lambda_user_value)
    )
  ) +
  labs(
    title = "Required Labeled Sample Size vs. Effect Size δ",
    subtitle = paste("N =", N, " unlabeled samples"),
    y = "Required labeled sample size n",
    x = "Effect size δ",
    color = "λ-mode"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    plot.title = element_text(face = "bold")
  )
```

## Summary

This vignette demonstrated:

- How to compute PPI / PPI++ sample size for mean estimation and regression.
- How to use direct moments or prediction metrics (continuous or classification).
- How to incorporate OLS or GLM models into contrast-based inference.
- When sample sizes become infeasible and how to handle caps.
- How $\lambda$ affects the required labeled sample size.

PPI++ typically yields the smallest labeled sample size, especially when the prediction model is strong and the correlation ( $\mathrm{Cov}(Y,f(X))$ ) is high.