---
title: "Prediction-Powered Power Planning"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Prediction-Powered Power Planning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup-chunk, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  fig.path = "man/figures/intro-ppi-"
)
```

## Overview

`pppower` provides analytical and Monte Carlo tools for **prediction-powered inference (PPI)**. We cover two main workflows:

* **Prediction-powered mean estimator** - the target parameter is the population mean \(\theta = \mathbb{E}[Y]\) corrected via predictions \(f(X)\).
* **Prediction-powered linear regression (PPI-OLS)** - the estimand is a linear contrast \(c^\top \beta\) from a PPI-adjusted OLS estimator.

Both workflows rely on a synthetic “superpopulation” that mimics the joint distribution of \((X, Y)\) and their cross-fitted predictions. The helper `simulate_crossfit_data()` generates such a data set.

```{r setup}
library(pppower)
set.seed(123)

df <- pppower:::simulate_crossfit_data(
  n = 4000,
  p = 5,
  family = gaussian(),
  K = 5
)
str(df)
```

---

## 1. Prediction-Powered Mean Estimation

### Notation

We test \(H_0: \theta = \theta_0\) versus a two-sided alternative, with:

- \(\delta = \theta - \theta_0\) - effect size of interest.
- \(N\) - unlabeled sample size used to evaluate predictions \(f(X)\).
- \(n\) - labeled sample size used to correct prediction error.
- \(\sigma_f^2 = \mathrm{Var}(f(X))\) and \(\sigma_{\text{res}}^2 = \mathrm{Var}(Y - f(X))\).

These can be estimated from the superpopulation.

```{r mean-inputs}
theta_true <- mean(df$y)
theta0 <- theta_true - 0.45
delta <- theta_true - theta0
var_f <- var(df$fhat_cf)
var_res <- var(df$y - df$fhat_cf)
N <- 2000
n <- 200
alpha <- 0.05

sim <- 2000
```

### Analytical power

`power_ppi_mean()` implements the normal-theory power formula, using the standard error \(\sqrt{\sigma_f^2 / N + \sigma_{\text{res}}^2 / n}\).

```{r mean-analytic}
power_ppi_mean(delta, var_f, var_res, N, n, alpha)
```

### Monte Carlo calibration (parametric)

`simulate_power()` mimics the null distribution with Monte Carlo samples of the test statistic.

```{r mean-simulate}
simulate_power(delta, var_f, var_res, N, n, alpha, R = sim)
```

### Monte Carlo calibration (resampling superpopulation)

`pppower:::simulate_power_ppi_mean()` repeatedly resamples labeled/unlabeled splits from the synthetic population, recomputing the PPI estimator each time.

```{r mean-pp}
mc_pp <- pppower:::simulate_power_ppi_mean(
  df = df,
  N = N,
  n = n,
  alpha = alpha,
  R = sim,
  theta0 = theta0,
  seed = 2024
)
mc_pp$empirical_power
mc_pp$analytical_power
```

### Sample-size planning

`n_required_PP()` solves for the labeled sample size that meets a power target.

```{r mean-n}
n_required_PP(
  delta = delta,
  N = N,
  alpha = alpha,
  power = 0.80,
  type = "mean",
  var_f = var_f,
  var_res = var_res
)
```

### Power and Type I error curves

`power_curve_mean()` sweeps over candidate labeled sample sizes, while `plot_type1_error_curve()` draws empirical vs. analytical rejection probabilities.

```{r mean-power-curve}
curve_fixed <- pppower:::power_curve_mean(
  n_grid = seq(50, 250, by = 50),
  delta = delta,
  N = N,
  var_f = var_f,
  var_res = var_res,
  alpha = alpha,
  R = sim
)

curve_fixed

plot(
  curve_fixed$n,
  curve_fixed$power_empirical,
  type = "b",
  pch = 19,
  col = "steelblue",
  xlab = "Labeled sample size n",
  ylab = "Power",
  main = "Power curve (mean estimator)"
)
lines(curve_fixed$n, curve_fixed$power_exact, col = "firebrick", lwd = 2, lty = 2)
legend("bottomright", legend = c("Empirical", "Analytical"),
       col = c("steelblue", "firebrick"), lty = c(1, 2), pch = c(19, NA), lwd = c(1, 2), bty = "n")
```

To study Type I error and power as a function of \(\delta\), use `type1_error_curve_mean()` or, for a fully simulated DGP, `type1_error_curve_mean_dgp()`:

```{r mean-type1}
effect_grid <- seq(-0.3, 0.3, by = 0.05)
curve_dgp <- pppower:::type1_error_curve_mean_dgp(
  effect_grid = effect_grid,
  N = N,
  n = n,
  family = gaussian(),
  superpop_n = 20000,
  p = 5,
  K = 5,
  alpha = alpha,
  R = sim
)

curve_dgp

pppower:::plot_type1_error_curve(
  curve_dgp,
  main = "Type I error / power (DGP)"
)
```

---

## 2. Prediction-Powered OLS (PPI-OLS)

### Notation

Let \(\beta\) be the PPI-adjusted coefficient vector and \(c\) the contrast of interest.

- Estimand: \(c^\top \beta\).
- Variance: \(c^\top (V_u / N + V_l / n) c\), where `V_u`/`V_l` are per-observation sandwich covariances for the unlabeled (imputation) and labeled (rectifier) regressions.

### Fit the model

```{r ols-fit}
fit_ols <- ppi_ols_fit(
  y = "y",
  f_col = "fhat_cf",
  formula = ~ x1 + x2 + x3 + x4 + x5,
  data_l = df[1:n, ],
  data_u = df[(n + 1):(n + N), ]
)

summary(fit_ols)
pieces <- fit_ols$pieces
```

`summary()` reports coefficient, standard error, and z-statistic. The `pieces` list exposes `V_u`, `V_l`, `n`, and `N` for power formulas.

### Construct a contrast

Suppose we focus on `x1`:

```{r ols-contrast}
coef_names <- colnames(pieces$V_u)
c_vec <- numeric(length(coef_names))
names(c_vec) <- coef_names
c_vec["x1"] <- 1

delta_ols <- sum(c_vec * fit_ols$coef)
delta_ols
```

`ppi_ols_wald()` provides the Wald test summary:

```{r ols-wald}
wald_x1 <- pppower:::ppi_ols_wald(fit_ols, A = matrix(c_vec, nrow = 1))
wald_x1
```

### Analytical power

```{r ols-analytic}
power_ppi_ols(
  delta = delta_ols,
  V_u = pieces$V_u,
  V_l = pieces$V_l,
  N = pieces$N,
  n = pieces$n,
  c = c_vec,
  alpha = alpha
)
```

### Monte Carlo calibration

`simulate_power_ppi_ols()` resamples from the superpopulation and recomputes the PPI-OLS estimator. Align the null value so the effect size matches the analytic call.

```{r ols-mc}
theta0 <- sum(c_vec * fit_ols$coef) - delta_ols

pppower:::simulate_power_ppi_ols(
  df = df,
  formula = ~ x1 + x2 + x3 + x4 + x5,
  N = pieces$N,
  n = pieces$n,
  c = c_vec,
  theta0 = theta0,
  alpha = alpha,
  R = sim,
  seed = 4321
)$analytical_power

pppower:::simulate_power_ppi_ols(
  df = df,
  formula = ~ x1 + x2 + x3 + x4 + x5,
  N = pieces$N,
  n = pieces$n,
  c = c_vec,
  theta0 = theta0,
  alpha = alpha,
  R = sim,
  seed = 4321
)$empirical_power
```

### Sample-size solver

```{r ols-n}
n_required_PP(
  delta = delta_ols,
  N = pieces$N,
  alpha = alpha,
  power = 0.80,
  type = "ols",
  V_u = pieces$V_u,
  V_l = pieces$V_l,
  c = c_vec
)
```

---

## Summary Table

| Task                                          | Functions                                                            |
|-----------------------------------------------|-----------------------------------------------------------------------|
| Generate cross-fitted superpopulation         | `simulate_crossfit_data()`                                           |
| Mean estimator power & calibration            | `power_ppi_mean()`, `simulate_power()`, `pppower:::simulate_power_ppi_mean()` |
| Mean estimator sample size                    | `n_required_PP(type = "mean", ...)`                                   |
| Mean estimator curves                         | `power_curve_mean()`, `type1_error_curve_mean[_dgp]()`, `plot_type1_error_curve()` |
| Fit prediction-powered OLS                    | `ppi_ols_fit()`, `summary()`                                          |
| OLS contrasts and power                       | `pppower:::ppi_ols_wald()`, `power_ppi_ols()`, `pppower:::simulate_power_ppi_ols()`       |
| OLS sample size                               | `n_required_PP(type = "ols", ...)`                                    |

Together these tools help you explore prediction quality, diagnose finite-sample effects, and plan labeled sample sizes for both mean and regression targets.
